---
title: "Mortality in South Africa"
runtime: shiny
output: 
  flexdashboard::flex_dashboard:
    theme: lumen
    orientation: columns
    vertical_layout: fill
    source_code: embed
---

```{r setup, include=FALSE}
library(flexdashboard)
library(southafricastats)
library(dplyr)
library(tidyr)
library(leaflet)
library(ggplot2)
library(DT)

totals = population_zaf %>% 
  filter(year == 2013) %>% 
  select(province, total)

compare_provinces = mortality_zaf %>% 
  left_join(totals) %>% 
  mutate(rate = deaths / total * 1e3) %>% 
  group_by(province, indicator) %>% 
  summarise(rate = mean(rate)) %>% 
  ungroup %>%
  left_join(totals) %>% 
  spread(indicator, rate)

mortality_zaf = mortality_zaf %>% 
  filter(indicator != 'All causes')

```


Column {.sidebar}
=============================================================

This app explores mortality data from the 
[South Africa Data Portal](http://southafrica.opendataforafrica.org/).

Use the input below to select a cause of death to explore.

```{r}
selectInput(inputId = 'indicator', 
            label = 'Choose a cause of death:', 
            choice = unique(mortality_zaf$indicator), 
            selected = 'Influenza and pneumonia (J10-J18)')
```

Map {data-icon="fa-map-marker"}
=======================================

### Where is the mortality rate highest?

```{r}
selected_df = reactive({
  subset_df = compare_provinces[ , c(1, which(colnames(compare_provinces) == input$indicator))]
  colnames(subset_df) = c('province', 'indicator')
  population_zaf %>% filter(year == 2013) %>% 
    left_join(subset_df) %>% 
    mutate(indicator = indicator / sum(indicator, na.rm = TRUE))
})

renderLeaflet({
  leaflet(selected_df()) %>% 
    addProviderTiles('CartoDB.Positron') %>% 
    addCircles(lng = ~longitude, 
               lat = ~latitude, 
               radius = ~sqrt(indicator) * 3e5, 
               weight = 2.5, 
               color = 'magenta', 
               popup = ~province)
})

```

Comparing provinces{data-icon="fa-list"}
=================================================

### How does the selected mortality rate compare to the overall mortality rate?

```{r}

scatterplot_df = reactive({
  subset_df = compare_provinces[ , c(1:2, 
                                     which(colnames(compare_provinces) == "All causes"), 
                                      which(colnames(compare_provinces) == input$indicator))]
  colnames(subset_df) = c('province', 'total', 'x', 'y')
  
  subset_df
})

renderPlot({
  ggplot(scatterplot_df(), 
       aes(x, 
           y, 
           label = province)) + 
          geom_point(aes(size = total), 
                     alpha = 0.8, color = 'midnightblue') + 
          geom_text(vjust = 2.5) + 
          theme_minimal() + 
    labs(x = 'All causes', y = input$indicator)
})

```


Changes in time {data-icon="fa-area-chart"}
===================================================

###How have the number of deaths changed over time?

```{r}
renderPlot({
  mortality_zaf %>% 
    filter(indicator == input$indicator) %>%
    ggplot(aes(year, deaths, color = province)) + 
    geom_line(alpha = 0.8, size = 1.5) + 
    theme_minimal(base_size = 18) + 
    labs(x = NULL, y = 'Number of deaths')
})
```

Tables {data-icon="fa-table"}
==============================================

### Check out the data here

```{r}
renderDataTable({
  datatable( mortality_zaf %>% 
    filter(indicator == input$indicator) 
    # %>%
    # ggplot(aes(year, deaths, color = province)) + 
    # geom_line(alpha = 0.8, size = 1.5) + 
    # theme_minimal(base_size = 18) + 
    # labs(x = NULL, y = 'Number of deaths')
    )
})
```


